---
title: "Cyclistic"
author: "Shubham"
date: '2022-06-11'
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Environment setup
Loading required packages
```{r}
library(tidyverse)
library(skimr)
library(lubridate)
library(dplyr)
library(scales)
```

### Loading datasets
```{r}
df1 <- read_csv("202106-divvy-tripdata.csv")
df2 <- read_csv("202107-divvy-tripdata.csv")
df3 <- read_csv("202108-divvy-tripdata.csv")
df4 <- read_csv("202109-divvy-tripdata.csv")
df5 <- read_csv("202110-divvy-tripdata.csv")
df6 <- read_csv("202111-divvy-tripdata.csv")
df7 <- read_csv("202112-divvy-tripdata.csv")
df8 <- read_csv("202201-divvy-tripdata.csv")
df9 <- read_csv("202202-divvy-tripdata.csv")
df10 <- read_csv("202203-divvy-tripdata.csv")
df11 <- read_csv("202204-divvy-tripdata.csv")
df12 <- read_csv("202205-divvy-tripdata.csv")
```

### Combining the datasets into one dataset
```{r}
trips <- bind_rows(df1, df2, df3, df4, df5,df6, df7, df8, df9, df10, df11, df12)

glimpse(trips)
summary(trips)
```

### Data Processing
Dropping all NA values
```{r}
trips_clean <- drop_na(trips)
```

Seperating the dates into day, month, year and day of the week
```{r}
trips_clean$date <- as.Date(trips_clean$started_at)
trips_clean$month <- format(as.Date(trips_clean$date), "%m")
trips_clean$year <- format(as.Date(trips_clean$date), "%Y")
trips_clean$day <- format(as.Date(trips_clean$date), "%d")
trips_clean$day_of_week <- format(as.Date(trips_clean$date), "%A")
```

creating custom columns of duration of the trips and total distance travelled
```{r}
trips_clean$trip_duration <- difftime(trips_clean$ended_at, trips_clean$started_at)

trips_clean$trip_duration_new <- abs(trips_clean$trip_duration)
```

Number of rides by usertype and its Visualization
```{r}
ridership <- trips_clean %>% 
  group_by(member_casual) %>% 
  summarize(count = n())

 ggplot(ridership) +
  geom_bar(aes(member_casual, y = count), position = "dodge", stat = "identity") +
   scale_y_continuous(labels = comma) +
   labs(title = "Count by Usertype", x = "Usertype")
```

Rides by months and its visualization
```{r}
rides_monthly <- trips_clean %>% 
  group_by(month = month(started_at , label = TRUE), member_casual) %>% 
  summarize(count = n(), groups = "drop")

rides_monthly %>% 
  ggplot(aes(x = month, y = count, fill = member_casual)) +
  geom_bar(poistion = "dodge", stat = "identity") +
  scale_y_continuous(label = comma) +
  labs(title = "Total rides by month", fill = "Ridership", x = NULL, y = "Total rides") 
```

Rides by week and its visualization
```{r}
rides_weekly <- trips_clean %>% 
  group_by(day_of_week, member_casual) %>% 
  summarize(count = n(), groups = "drop") %>% 
  arrange(match(day_of_week,c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")))

rides_weekly$day_of_week <- fct_inorder(rides_weekly$day_of_week)
rides_weekly %>% 
  ggplot(aes(x = day_of_week, y = count, fill = member_casual)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_y_continuous(label = comma) +
  labs(title = "Total rides by days of the week", fill = "Ridership", x = NULL, y = "Total rides")
```

### Saving and exporting files for furthur detailed visualization in Tableau
```{r}
write.csv(ridership, file = "cyclistic-ridership.csv")
write.csv(rides_weekly,file = "cyclistic-rides-of-week.csv")
write.csv(rides_monthly,file ="cyclistic-rides-of-month.csv")
write.csv(trips_clean,file ="cyclistic-trips-data.csv")

```

